---
title: "Trip data analytics"
output: html_notebook
---

Divvy_Exercise_Full_Year_Analysis

This analysis is based on the Divvy case study "'Sophisticated, Clear, and Polished': Divvy and Data Visualization" written by Kevin Hartman (found here: <https://artscience.blog/home/divvy-dataviz-case-study>). The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: "In what ways do members and casual riders use Divvy bikes differently?"

Install required packages

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
getwd()
```

Read trip data set from csv
```{r}
tripdata_4 <- read_csv("202004-divvy-tripdata.csv")
tripdata_5 <- read_csv("202005-divvy-tripdata.csv")
tripdata_6 <- read_csv("202006-divvy-tripdata.csv")
```

Check column name
```{r}
colnames(tripdata_4)
colnames(tripdata_5)
colnames(tripdata_6)
```

Check Str
```{r}
str(tripdata_4)
str(tripdata_5)
str(tripdata_6)
```


Combine into single file
```{r}
all_trips <- bind_rows(tripdata_4, tripdata_5, tripdata_6)
all_trips <- all_trips %>%
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

Inspect the combined table
```{r}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)
```

Extract date, month, day, year
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
Add ride_length
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
str(all_trips)
```

Convert ride_length to numeric
```{r}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

Delete rows which have nagative ride_length
```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length < 0),]
```

Calc mean, median, max, min of ride_length
```{r}
summary(all_trips_v2$ride_length)
```


Compare members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

See the average ride time by each day for members vs casuals
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

fix day of week
```{r}
#all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=1:7, labels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
str(all_trips_v2)
```

calculate the average ride time by each day for members vs casual again
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

analyze ridership data by type and weekday
```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarize(number_of_ride = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```

Visualize the number of rises by rider type
```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarize(number_of_ride = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_ride, fill = member_casual)) +
  geom_col(position = "dodge")
  
```

Visualize average_duration by rider_type
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

Export summary file to csv
```{r}
counts <- all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarize(number_of_ride = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
write.csv(counts, file = 'tripdate_output_summary.csv')
```


